---
title: "Reliability analysis"
author: "Sami Muraja & Maria Kokkinen"
date: "2023-10-20"
output: html_document
---

# Reliability Analysis

Load data:
```{r eval = FALSE}
load("wrangled-data.RData")
library(tidyverse)
library(psych)
```

```{r, eval = FALSE}
# to see the results without the around.G reversed, run this before the rest
DJGLS[, "around.G"] = 6 - DJGLS[, "around.G"]
DJGLS.[, "around.G"] = 6 - DJGLS.[, "around.G"]
```

## Testing the assumptions of the tau-equivalent model

### Unidimensionality & Tau-equivalence

#### DJGLS
Making a scree plot:
```{r}
# pca
results <- prcomp(DJGLS_det[-1])
results$rotation %>% as.data.frame

# calculate total variance explained by each principal component
var_explained = results$sdev^2 / sum(results$sdev^2)

#create scree plot
qplot(c(1:6),var_explained) + 
  geom_line() + 
  xlab("Principal Component") + 
  ylab("Variance Explained") +
  ggtitle("Scree Plot G") +
  ylim(0, 1)
```


```{r}
library(psych)
fa(DJGLS_det[-1], nfactors = 2, rotate = "oblimin")
fa(DJGLS_det[-1], nfactors = 2, rotate = "varimax")
```


```{r}
fa(DJGLS_det[-1], nfactors = 1, rotate = "oblimin")
fa(DJGLS_det[-1], nfactors = 3, rotate = "oblimin")
```


```{r}
library(psych)
fa(DJGLS_mean[-1], nfactors = 2)
fa(DJGLS_det[-1], nfactors = 2)
fa(DJGLS_sto[-1], nfactors = 2)
```
The imputation method has a negligible influence on the factor analysis results with two factors.

```{r}
fa(DJGLS_det[-1], nfactors = 1)
fa(DJGLS_det[-1], nfactors = 3)
```
One factor is evidently too little: the communalities in the emotional loneliness factor are all very low.

#### ALS

```{r}
# In the 6-item G validation study they used pca, so checking how that works for this data
r <- CSNe_det %>% 
  select(-ID) %>% 
  cor(use = "complete.obs")

# with scree plot
results <- prcomp(CSNe_det[-1])
results$rotation %>% as.data.frame

# scree plot
#calculate total variance explained by each principal component
var_explained = results$sdev^2 / sum(results$sdev^2)

#create scree plot
qplot(c(1:4),var_explained) + 
  geom_line() + 
  xlab("Principal Component") + 
  ylab("Variance Explained") +
  ggtitle("Scree Plot DJGLS") +
  ylim(0, 1)
```

```{r}
library(psych)
# allardt, CSNA vs CSNe
fa(CSNA[-1], nfactors = 1)
#fa(A[-1], nfactors = 2)
#fa(A[-1], nfactors = 3)

fa(CSNe[-1], nfactors = 1)
#fa(CSNe[2:5], nfactors = 2)
#fa(CSNe[2:5], nfactors = 3)

fa(CSNe_det[-1], 1) # imputation makes negligible difference
```

### Uncorrelated error terms

```{r}
library(lavaan)
###
scale <- 'ELF =~ empty.G + around.G + reject.G
         SLF =~ rely.G + trust.G + close.G'
Gfit <- cfa(scale, data=DJGLS_det[-1], std.lv=TRUE, auto.cov.lv.x=FALSE)
residuals(Gfit)$cov

###
scale  <- 'f  =~ interest.A + happy.A + easy.A + lonely.A'
CSNefit <- cfa(scale, data=CSNe_det[-1]) 
residuals(CSNefit)$cov
```

## Reliability analysis with imputation sensitivity testing

```{r}
# loading scripts compiled by Reijo Sund. They can be downloaded from https://github.com/rsund/reliab
#source("factor_score_reliability.R")
source("reli-fun.R")

# loading an extension of these functions that includes preparation of the data and imputation sensitivity testing
source("relisens.R")
```

### DJGLS

```{r}
relisens(DJGLS[-1], 2, method = "score")
```

```{r}
relisens(DJGLS[-1], 2, method = "sum")
```

```{r}
relisens(DJGLS[-1], 2, method = "image")
```

```{r}
source("crelisens.R")
crelisens(DJGLS[-1], items.until = 3)
```

#### DJGLS per factor

##### SLF - Social loneliness factor

```{r}
SLF <- DJGLS. %>% select(rely.G, trust.G, close.G)
SLF <- SLF[!rowSums(is.na(SLF)) > 1,]

relisens(SLF, 1, method = "score") 
relisens(SLF, 1, method = "sum") 
relisens(SLF, 1, method = "image") 
crelisens(SLF, items.until = 3)
```

##### ELF - Emotional loneliness factor

```{r}
ELF <- DJGLS. %>% select(empty.G, around.G, reject.G)
ELF <- ELF[!rowSums(is.na(ELF)) > 1,]

relisens(ELF, 1, method = "score") 
relisens(ELF, 1, method = "sum") 
relisens(ELF, 1, method = "image") 
crelisens(ELF, items.until = 3)
```

### ALS

```{r}
relisens(CSNe[-1], 1, method = "score")
```

```{r}
relisens(CSNe[-1], 1, method = "sum")
```

```{r}
relisens(CSNe[-1], 1, method = "image")
```

```{r}
crelisens(CSNe[-1], 4)
```


#### Influence of CS as neutral (CSNe) versus CS as missing value (CSNA)

```{r}
CS_diff <- function(reliNe, reliNA){
  diffNANe <- reliNA %>% as.data.frame()
  #diff[3:10] <- reliNALS[3:10] - reliNe[3:10]
  diffNANe[3:6] <- reliNe[3:6] - reliNA[3:6]
  diffNANe[3:6] <- diffNANe[3:6] %>% format(digits = 2)
  diffNANe
}
# factor scores
reliNA <- CSNA[-1] %>% relisens(1, round = F)
reliNe <- CSNe[-1] %>% relisens(1, round = F)
CS_diff(reliNe, reliNA)

# unweighted sums
reliNA <- CSNA[-1] %>% relisens(1, round = F, method = "sum")
reliNe <- CSNe[-1] %>% relisens(1, round = F, method = "sum")
CS_diff(reliNe, reliNA)

# factor images
reliNA <- CSNA[-1] %>% relisens(1, round = F, method = "image")
reliNe <- CSNe[-1] %>% relisens(1, round = F, method = "image")
CS_diff(reliNe, reliNA)
```
